Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-09-09 summary.Rmd
typo
minor tweak to clustering function

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$. Note that $P^{(kl)} = (P^{(lk)})^\top$.  
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. As in the 
$k = l$ case, we note that the singular value decomposition 
$P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. (We can also note that the SVD of 
$P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top$, i.e., 
$\sigma_{kl} = \sigma_{lk}$, $u^{(kl)} = v^{(lk)}$, and 
$v^{(kl)} = u^{(lk)}$.)  
Now consider the Hermitian dilation 
$M^{(kl)} = 2 \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$, 
which is a symmetric $(n_k + n_l) \times (n_k + n_l)$ matrix. It can be shown 
that the eigendecomposition of $M^{(kl)}$ is  
$M^{(kl)} = 
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix}^\top$.  
Thus treating $M^{(kl)}$ as the edge probability matrix of a GRDPG, we have 
latent positions in $\mathbb{R}^2$ given by 
$\begin{bmatrix} 
  \sigma_{kl} u^{(kl)} & \sigma_{kl} u^{(kl)} \\ 
  \sigma_{kl} v^{(kl)} & -\sigma_{kl} v^{(kl)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(kl)} & \lambda^{(kl)} \\ 
  \lambda^{(lk)} & -\lambda^{(lk)} 
\end{bmatrix}$.  
Now consider $M' = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$. 
Then $M' = M + E'$ where 
$E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}$ and $E$ is the 
$n_k \times n_l$ matrix of independent noise (to generate the Bernoulli entries 
in $A^{(kl)}$.  
Then $M'$ is an adjacency matrix drawn from $M^{(kl)}$, so its adjacency 
spectral embedding, given by 
$\begin{bmatrix} 
  \hat{\lambda}^{(kl)} & \hat{\lambda}^{(kl)} \\ 
  \hat{\lambda}^{(lk)} & -\hat{\lambda}^{(lk)} 
\end{bmatrix}$ 
where each $\hat{\lambda}^{(kl)}$ is defined as in Algorithm 4, approximates 
the latent positions of $M^{(kl)}$ by the rate in Theorem 5 of 
\citet{rubindelanchy2017statistical}.  
In this case, the indefinite orthogonal transformation $W_*$ in the GRDPG 
result \cite{rubindelanchy2017statistical} is of the form 
$\hat{U}^\top \hat{U}$. The eigenvalues of $M$ are distinct since the signature 
for this GRDPG is $(1, 1)$, and $\hat{U}^\top \hat{U}$ is block diagonal, 
resulting in $W_* = I$. Therefore, the adjacency spectral embedding of $M'$ is 
a direct estimation of the specific latent positions outlined for $M^{(kl)}$.

@@

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}. The original data consists of 419 vertices, 
each corresponding to a British MP belonging to one of five political parties. 
For this data analysis, we subsetted the data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
dataset, resulting in 329 vertices and two parties ($n = 329, K = 2$). 
\citet{307cbeb9b1be48299388437423d94bf1} were able to obtain a classification 
accuracy of `r round(328 / 329 * 100, 1)`\%. Algorithm 2 obtained a lower 
classification accuracy of `r round(clust.acc * 100, 1)`\%.

>>>>

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$. Note that $P^{(kl)} = (P^{(lk)})^\top$.  
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. As in the 
$k = l$ case, we note that the singular value decomposition 
$P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. (We can also note that the SVD of 
$P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top$, i.e., 
$\sigma_{kl} = \sigma_{lk}$, $u^{(kl)} = v^{(lk)}$, and 
$v^{(kl)} = u^{(lk)}$.)  
Now consider the Hermitian dilation 
$M^{(kl)} = 2 \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$, 
which is a symmetric $(n_k + n_l) \times (n_k + n_l)$ matrix. It can be shown 
that the eigendecomposition of $M^{(kl)}$ is  
$M^{(kl)} = 
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix}^\top$.  
Thus treating $M^{(kl)}$ as the edge probability matrix of a GRDPG, we have 
latent positions in $\mathbb{R}^2$ given by 
$\begin{bmatrix} 
  \sigma_{kl} u^{(kl)} & \sigma_{kl} u^{(kl)} \\ 
  \sigma_{kl} v^{(kl)} & -\sigma_{kl} v^{(kl)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(kl)} & \lambda^{(kl)} \\ 
  \lambda^{(lk)} & -\lambda^{(lk)} 
\end{bmatrix}$.  
Now consider $M' = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$. 
Then $M' = M + E'$ where 
$E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}$ and $E$ is the 
$n_k \times n_l$ matrix of independent noise (to generate the Bernoulli entries 
in $A^{(kl)}$.  
Then $M'$ is an adjacency matrix drawn from $M^{(kl)}$, so its adjacency 
spectral embedding, given by 
$\begin{bmatrix} 
  \hat{\lambda}^{(kl)} & \hat{\lambda}^{(kl)} \\ 
  \hat{\lambda}^{(lk)} & -\hat{\lambda}^{(lk)} 
\end{bmatrix}$ 
where each $\hat{\lambda}^{(kl)}$ is defined as in Algorithm 4, approximates 
the latent positions of $M^{(kl)}$ by the rate in Theorem 5 of 
\citet{rubindelanchy2017statistical}.  
In this case, the indefinite orthogonal transformation $W_*$ in the GRDPG 
result \cite{rubindelanchy2017statistical} is of the form 
$\hat{U}^\top \hat{U}$. The eigenvalues of $M$ are distinct since the signature 
for this GRDPG is $(1, 1)$, and $\hat{U}^\top \hat{U}$ is block diagonal, 
resulting in $W_* = I$. Therefore, the adjacency spectral embedding of $M'$ is 
a direct estimation of the specific latent positions outlined for $M^{(kl)}$, 
up to sign flip.

@@

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}. The original data consists of 419 vertices, 
each corresponding to a British MP belonging to one of five political parties. 
For this data analysis, we subsetted the data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
dataset, resulting in 329 vertices and two parties ($n = 329, K = 2$). 
\citet{307cbeb9b1be48299388437423d94bf1} were able to obtain a classification 
accuracy of `r round(328 / 329 * 100, 1)`\% via modularity maximization. 
Algorithm 2 obtained a lower 
classification accuracy of `r round(clust.acc * 100, 1)`\%. We can also note 
that Algorithm 2 had a much shorter runtime than modularity maximization.
